---
title: "xglue"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warn=FALSE)
library(xglue)
```

## Motivation for xglue

I wrote xglue because for a scientific paper I needed to generate several customized Latex tables that each contained the results of many statistical tests. These were not standardized tables that could be quickly generated e.g. with [stargazer](https://cran.r-project.org/web/packages/stargazer/index.html). 
There existed Latex code for a similar table, but who wants to update a lot of numbers by hand? Building the table by pasting R strings together was also inconvenient. E.g. Latex has many backslashes \\, which must be converted to double backslashes \\\\ inside R strings making things even harder to read.

xglue solves this problem by having templates as files that include placeholders and blocks that allow to collapse vectorized expressions or perform group_by operation on data frames to combine strings.

## Starting example: A template with collapse block

Here is a simple example of an xglue template:

```
Hello 
#< collapse ","
{names}
#>
!

Greetings, {myname}
```

The following code loads the template and then fills the fields `{name}` and `{myname}` with the values of the corresponding R variables.
```{r}
library(xglue)
tpl = readLines(system.file("examples/hello_tpl.txt",package="xglue"))
names = c("Anton","Berta","Charly")
myname = "Sebastian"
xglue(tpl)
```

The main difference to the [glue](https://cran.r-project.org/web/packages/glue/index.html) package is that our template can contain blocks, like the collapse block.

```
#< collapse ","
{names}
#>
```
It means that the elements of the vector `names` will be collapsed to a single string as if calling in R `paste(names, collapse=",")`.

Special blocks always open in a new line starting with `#<` and close with a `#>` line. Note that in the created content there will no new line before and after the block i.e. we get the output
```
Hello Anton, Berta, Charly!
```
instead of
```
Hello
Anton, Berta, Charly
!
```
If you want a line break before or after a block just add an empty line before or after it.


## Example with use data block

Consider the following subset of the `starwars` data set in the `dplyr` package.

```{r}
library(dplyr)
data(starwars)
dat = starwars %>%
  filter(species %in% c("Droid","Wookiee")) %>%
  select(species, name, homeworld)
dat
```

Assume we want to generate a Markdown file that has a subsection for each species which contains a table of the characters of that species. The result should look as follows:

```
## Star Wars Characters By Species

Number of total characters: 8

### Characters of species Droid (6)

|Name                |Homeworld           |
|--------------------|--------------------|
|C-3PO               |Tatooine            |
|R2-D2               |Naboo               |
|R5-D4               |Tatooine            |
|IG-88               |NA                  |
|R4-P17              |NA                  |
|BB8                 |NA                  |

### Characters of species Wookiee (2)

|Name                |Homeworld           |
|--------------------|--------------------|
|Chewbacca           |Kashyyyk            |
|Tarfful             |Kashyyyk            |
```

Here is the xglue template we use:

```
## Star Wars Characters By Species

Number of total characters: {nrow(dat)}

#< use dat; group_by species; collapse "<<newline>>"

### Characters of species {first(species)} ({nrow(.DATA)})

#< ignore
The created markdown tables will look e.g. as follows

|Name                |Homeworld           |
|--------------------|--------------------|
|Chewbacca           |Kashyyyk            |
|Tarfful             |Kashyyyk            |

#>
|{fill('Name',20,'')}|{fill('Homeworld',20,'')}|
|{fill('',20,'-')}|{fill('',20,'-')}|

#< collapse <<newline>>
|{fill(name,20,' ')}|{fill(homeworld),20,' ')}|
#>
#>
```

We now have a use data block starting with the line 

```
#< use dat; group_by species; collapse "<<newline>>"
```
It means that we take the data frame `dat` from the R environment. We then split it into one sub data frame per `species` and apply the template inside the block for each sub data frame and collapse the results to a single string adding a line break between each chunk. (The `<<newline>>` is a special symbol for the collapse argument. You can change it with the `newline` argument in the call to `xglue`.) 

The template inside the block should return a single string for each sub data frame. It is similar to a dplyr `summarize` command that should return single value for each group.

For this reason, we e.g. use the function `first` in our field `{first(species)}` for the subsection header. Inside the block, you can access the special variable `.DATA` that contains the current sub data frame, see e.g. the `{nrow(.DATA)}` field.

There are other blocks nested inside the use block. An `#< ignore` block can be used to comment your template. Everything inside that block will be ignored.

More importantly, the collapse block
```
#< collapse <<newline>>
|{fill(name,20,' ')}|{fill(homeworld),20,' ')}|
#>
```
generates the rows of each table.

The function `fill` is a custom function that we specify before our call to `xglue`. Here is the code that then generates the markdown file from this template: 
```{r eval=FALSE}
library(xglue)
# A helper function used in the template
fill = function(str="",len=10, fill=" ") {
  add = paste0(rep(fill, len),collapse="")
  str=substring(paste0(str,add),1, len)
  str
}
tpl = readLines(system.file("examples/starwars_tpl.md",package="xglue"))
xglue(tpl)
```

## Adapting factors using the forcats packages

A powerful complement to the xglue package is the [forcats](https://cran.r-project.org/web/packages/forcats/index.html) that simplifies ordering and other manipluations of factor variables. Here is an example:
```{r eval=FALSE}
library(forcats)
dat = starwars %>%
  mutate(
    homeworld = fct_explicit_na(homeworld,"unknown"), 
    species = species %>%
      as.factor() %>%
      # Sort by frequency
      fct_infreq(TRUE) %>%
      # Lump together if there are less than 2 species
      fct_lump_min(2,other_level = "Other") %>%
      # Change NA value also to Other
      fct_explicit_na("Other")
  )%>%
  arrange(species, name)
xglue(tpl)
```

The result is shown [here](???).

## Using a function for more generic specification of table columns

In our example above, we fixed the two columns `Name` and `Homeworld` in the template. What about a more generic template where we can specify arbitrary table columns in our R code?

For the markdown table, the simplest way is to use an R function that generates such a table. Consider this template:

```
## Star Wars Characters

Number of total characters: {nrow(dat)}

#< use dat; group_by species; collapse <<newline>>

### Characters of species {first(species)} ({nrow(.DATA)})

{md_table(.DATA, cols)}
#>
```

We can use it with the following code:
```{r eval=FALSE}
md_table = function(data, cols) {
  str = knitr::kable(data[,cols], "pipe")
  str = paste0(str, collapse="\n")
}
cols = c("name","homeworld","hair_color")
tpl = readLines(system.file("examples/starwars2_tpl.md",package="xglue"))
xglue(tpl)
```

### Using pre blocks

An alternative using R functions for such more dynamic templates are `pre` blocks. Consider the following example that also generates our Star Wars Markdown tables with customly specified columns:

```
## Star Wars Characters By Species

#< use dat; group_by species; collapse <<newline>>

### Characters of species {first(species)} ({nrow(.DATA)})

#< pre
|
#< collapse
{fill('<<cols>>',<<ns>>,' ')}|
#>

|
#< collapse
{fill('',<<ns>>,'-')}|
#>
#>

#< collapse <<newline>>

#< pre
|
#< collapse
{fill(format.na(<<cols>>),<<ns>>,' ')}|
#>
#>
#>
#>
```
The code in the `pre` blocks will be first compiled. These blocks can generate the code for the actual templates. When the pre blocks are compiled, all fields inside a `<<` and `>>` block (can be customized with the arguments `pre.open` and `pre.close` to xglue) will be replaced. Other blocks, like `collapse` blocks will also be run inside a pre block.

We can just generate the resulting template by setting the argument `just.pre=TRUE`:

```{r eval=FALSE}
pre.tpl = readLines(system.file("examples/starwars_pre_tpl.md",package="xglue"))
tpl = xglue(pre.tpl)
```

The resulting template looks as follows

```
## Star Wars Characters By Species

#< use dat; group_by species; collapse <<newline>>

### Characters of species {first(species)} ({nrow(.DATA)})

|{fill('name',20,' ')}|{fill('homeworld',12,' ')}|{fill('hair_color',10,' ')}|
|{fill('',20,'-')}|{fill('',12,'-')}|{fill('',10,'-')}|

#< collapse <<newline>>
|{fill(format.na(name),20,' ')}|{fill(format.na(homeworld),12,' ')}|{fill(format.na(hair_color),10,' ')}|
#>
#>
```

If you keep the default `just.pre=FALSE` then `xglue` just returns the results of compiling this template again.

Addmittedly, templates with pre block may be not very easy to read, but still they can be helpful in certain instances.

## Installation

The package is hosted on [r-universe](https://skranz.r-universe.dev/ui#builds). To install it run:

```r
options(repos = c(
    skranz = 'https://skranz.r-universe.dev',
    CRAN = 'https://cloud.r-project.org'))
install.packages('xglue')
```

